Long- and Short-Read Metabarcoding Give Similar Results for Spatial Turnover Range in an Ectomycorrhizal Savanna Woodland

### 2020-02-18 writing/transect_paper.Rmd
clean up before sending to Mohammad and Nourou

knitr::opts_chunk$set(
  dev = "pdf",
  # dev.args = list(pointsize=32),
  echo = FALSE,
  # results = 'asis',
  # fig.show = 'asis',
  fig.width = 6,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  external = TRUE
)
options(
  tikzDefaultEngine = "luatex",
  tikzXelatexPackages = c(options("tikzXelatexPackages")$tikzXelatexPackages,
                          "\\usepackage[version=4]{mhchem}\n",
                          "\\usepackage{siunitx}\n",
                          "\\DeclareSIUnit\\year{a}\n",
                          "\\usepackage{fontspec}"),
  tikzMetricsDictionary = here("writing/transect_paper-tikzDictionary")
)

@@

Ectomycorrhizal (ECM) communities are characterized by mutualistic symbiosis between certain fungi and plant roots.
Spatial turnover rate is one important characteristic of ecological communities, which is relevant for biodiversity assessment and sampling design.
We investigated turnover rate in an ECM fungal community using soil metabarcoding with four different sequencing strategies: short amplicon sequencing of the ITS2 region (300--400 bp) with Illumina MiSeq, Ion Torrent Ion S5, and PacBio RS II, as well as long amplicon sequencing of the full ITS and partial LSU regions (1200--1600 bp) with PacBio RS II.
Resulting turnover rates depended more on statistical method than sequencing technology.
We found significant autocorrelation in distance classes <3 m using a Mantel correlogram, and a best-fit distance-decay curve range of 17--19 m, both shorter than previously reported for similar habitats.
The use of long-amplicon sequencing enables construction of a phylogenetic tree from metabarcoding reads, which facilitates taxonomic identification of sequences.
However, long reads present issues for denoising algorithms in diverse communities.
We present a solution that splits the reads into shorter homologous regions prior to denoising, and then reconstructs the full denoised reads.
In the choice between short and long amplicons, we suggest a hybrid approach using short amplicons for sampling breadth and depth, and long amplicons to characterize the local species pool for improved identification and phylogenetic analyses.

@@

Several studies have investigated the turnover scale in ECM forests, but so far they have been based on analysis of ECM root tips, rather than bulk soil.
@lilleskov2004 found autocorrelation only at ranges <2.6 m at most sites in temperate forests, based on Sanger sequencing.
Similarly, @pickles2012a found autocorrelation at distances <3.4 m, also in temperate forests, based on T-RFLP.
Two studies have included data from tropical African woodlands and wooded savannahs.
@tedersoo2011 found autocorrelation at ranges <10 m in Miombo woodland based on Sanger sequencing.
However, this study was based on irregular sampling with a minimum spacing of 8 m, so this as the smallest distance class.
@Bahram2013 performed a meta-analysis of turnover scale in ECM communities, including data from Benin, where they found turnover range of 65 m based on distance-decay curves.
They concluded that the range of distance decay is greater in tropical forests than temperate forests.

@@

Sampling was conducted at two sites (Ang: N 9.75456° W 2.14064°; Gan: N 9.75678° W 2.31058 °) approximately 30 km apart in the _Forêt Reservée de l'Ouémé Supérieur_ (Upper Ouémé Forest Reserve) in central Benin.
Both locations were located in woodlands dominated by the ECM host tree _Isoberlinia doka_ (Caesalpinioideae).
At each site, 25 soil samples were collected along a linear transect at intervals of 1 m in May 2015.
One third of the sample locations (3 m spacing) were resampled one year later in June 2016.
For each sample, any coarse organic debris <!--Do you have any definition? Wider than 5 mm? MR--> was removed from the soil surface and a sample of approximately 5cm×5cm×5cm was extracted with a <!-- sterilized? -MR--> knife blade.
Each sample was sealed in a plastic zipper bag and homogenized by shaking and manually breaking apart soil aggregations.
Approximately 50 mg total of soil was collected from two locations in the homogenized soil sample and placed into a separate 2.0 mL microtube containing 750 mL of lysis buffer and lysis beads (Xpedition^TM^ Soil/Fecal DNA miniprep, Zymo Research Corporation, Irvine, California, USA) and lysed in the field using a handheld bead-beater (TerraLyser^TM^; Zymo Research Corporation).
<!--The remainder of each sample was dried at 40°C with an electric dehydrator within 24-48 hours of collection for bulk chemical analysis. -->

@@

After field lysis, DNA was extracted using the Xpedition^TM^ Soil/Fecal Prep kit (see above).
Samples preserved using LifeGuard were first centrifuged at 10000×g for 1 minute, after which the supernatant was removed and DNA was extracted from the remaining soil using the Soil/Fecal Prep kit as for the other samples.
DNA was quantified using fluorometrically using Quant-iT^TM^ PicoGreen^TM^ dsDNA (Thermo Fisher Scientific, Waltham, MA, USA) fluorescent indicator dye on a Infinite F200 plate spectrofluorometer (Tecan Trading AG, Männedorf, Switzerland) according to the manufacturer's protocol.

Two different fragments of the nuclear rDNA were amplified (Figure\ \@ref(fig:rDNA)).
The short amplicon (approximately 300 bp) targetted the full ITS2 region as well as parts of the flanking 5.8S and large subunit (LSU) rDNA, using gITS7 [@ihrmark2012] as the forward primer and a mix of ITS4 [@white1990amplification] and ITS4a [@urbina2016] as the reverse primer (hereafter, ITS4m).
The long amplicon (approximately 1500 bp) targetted the full ITS region including the 5.8S rDNA and approximately 950 bp at the 5' end of the LSU, including the first three variable regions (Figure\ \@ref(fig:rDNA)), using ITS1 [@white1990amplification] as the forward primer and LR5 [@vilgalys1990] as the reverse primer.

The gITS7 primers for the short amplicon were indexed for multiplexing (**Supplementary File**).
Amplification was performed by polymerase chain reaction (PCR) in 20µl reactions containing 200 µM dNTP mix, 250 µM indexed gITS7 primer, 150µM ITS4m, 2mM MgCl~2~, 0.1 U *Taq* polymerase (Dream *Taq*, Thermo Fisher Scientific, Waltham, MA, USA) and 3--7 ng purified DNA in Dream *Taq* buffer.
The reaction conditions were 10 min at 95°, followed by 35 cycles of 60 s at 95°, 45 s at 56°, and 50 s at 72°, and finally 3 min at 72°.
Each reaction was conducted in three technical replicates to reduce the effect of PCR stochasticity, which were pooled after amplification.

@@

<!--If limited by space, this could be in supp, and then just described in text. -MR-->
```{tikz rDNA, fig.cap = "(ref:rDNA)", code = readLines("rDNA_amplicons.pgf"), fig.ext = "pdf"}

@@

Both primers for the long amplicon were indexed for multiplexing (**Supplementary File**).
PCR was performed as for the short amplicons, but with 500 µM of each of the two primers.
Reaction conditions were 10 min at 95°, 30 cycles of 45 s at 95°, 45 s at 59°, and 90 s at 72°, and finally 10 min at 72°.
Each reaction was performed in three technical replicates as for short amplicons.

@@

Amplicons were purified using SPRI beads [@vesterinen2016] and quantified fluorometrically as above.
An aliquot of 100 ng of DNA from each sample (or the total PCR product if less than 100 ng) was pooled into two libraries each for long and short amplicons.
Each library was sequenced using Single Molecule Real Time (SMRT) sequencing on a Pacific Biosciences (PacBio) RS II sequencer at the Uppsala Genome Center (UGC; Uppsala Genome Center, Science
for Life Laboratory, Dept. of Immunology, Genetics and Pathology, Uppsala University, BMC, Box 815, SE-752 37 UPPSALA, Sweden).
Short amplicon libraries were sequenced on two SMRT cells each, while long amplicon libraries were sequenced on four SMRT cells each.

Additionally the short amplicon libraries were combined and sequenced using an Ion S5  (Ion Torrent) sequencer using one 520 chip at UGC, and a MiSeq (Illumina Inc.) sequencer using v3 chemistry with a paired-end read length of 300 bp at the SNP&SEQ Technology Platform (Dept. of Medical Sciences, Uppsala University, BMC, Box 1432, SE-751 44 UPPSALA, Sweden).
The Illumina library was pooled with samples for another project, with one half of the reads from one lane devoted to this project.

@@

Raw reads were divided into shorter regions by matching to covariance models (CM), which are similar to stochastic hidden markov models (HMM), but account for both nucleotide sequence and RNA secondary structure [@eddy1994].
First, the 5.8S rDNA was located in each read by searching for Rfam model RF0002 [@kalvari2018] using `cmsearch` from Infernal 1.1.2 [@nawrocki2013], and all bases before the 5.8S were assigned to ITS1.
No attempt was made to remove the approximately 12 bp fragment of the SSU from the 5' end of ITS1 in the long amplicons; it was too short to be reliably detected by a CM or the HMMs employed by ITSx [@bengtsson-palme2013].
<!--This approach was able to identify both the 5' and 3' ends of the 5.8S rDNA more consistently and quickly than ITSx [@bengtsson-palme2013, data not shown].-->
A reference alignment including conserved RNA base pairing between and within the 5.8S and relevant portions of LSU was generated from the fungal 28S RNA seed alignment from the Ribosomal Data Project (RDP) release 11.5 [@glockner2017; @cole2014] by truncating after the LR5 primer site and using the reference line to annotate the variable regions *sensu* @michot1984 and @raue1988.
A CM was generated from the alignment using `cmbuild` from Infernal.
The fragment of each read beginning with the 5.8S rDNA was then aligned to the CM using `cmalign` from Infernal.
The annotation line in the CM alignment for each read was then used to split the reads into alternating more-conserved and less-conserved regions as shown in Figure\ \@ref(fig:rDNA),
where LSU1-4 represent the conserved regions of LSU flanking the variable D1-3 regions [@michot1984].
For short amplicons, only (partial) 5.8S, ITS2, and (partial) LSU1 were extracted.
Code to extract the regions, including annotated seed alignments and CMs, is available in the new R package `LSUx`, available on github (https://github.com/brendanf/LSUx).

@@

Unique LSU regions from long amplicon ASVs were aligned using DECIPHER [@wright2015] with up to 10 iterations of progressive alignment and conserved RNA secondary structure calculation and 10 refinement iterations.
<!-- The LSU tree ends up not looking as nice as the tree which includes ITS (I don't know why, except that maybe 5.8S is more helpful than ITS is harmful?) so I'm using that. Maybe I don't need to mention the LSU-only tree.-->
This alignment was truncated at a position after the D3 region corresponding to base 907 of the *Saccharomyces cerevisiae* S288C reference sequence for LSU, because several sequences had introns
<!-- Just curious, what are Type 1 introns -AR--> <!-- It has to do with the enzyme that cuts them out, I think, but I don't actually know my introns are Type 1, just that this is a location where Type 1 introns are known to occur. -BF-->
after this position, as also observed in several fungal species by @holst-jensen1999.
Full length long amplicon ASVs (including ITS1, 5.8S, and ITS2 regions) were aligned and truncated in the same way.

@@

Ecological community dissimilarity matrices were calculated using the ASV/OTU based Bray-Curtis metric (both long and short amplicons) and the phylogenetically based weighted UNIFRAC method (only long amplicons) in `phyloseq` version `r packageVersion("phyloseq")`.
Each of these distance matrices was used to calculate a Mantel correlogram for distances of 0--12 m.
Separate correlograms were drawn for samples taken during the same year, and samples separated in time by one year, in order to assess the degree to which the soil community changes over the course of one year.

@@

DNA concentrations after extraction and PCR, as well as sequencing reads for PacBio and Illumina, are shown per sample in Supplementary Figure \@ref(fig:sample_depth).
Samples from Ang in 2015 yielded low quantities of DNA, poor PCR performance, and ultimately very few sequencing reads, especially in the long amplicon library, where only one sample produced more than 100 reads.
Consequently Ang samples were excluded from spatial analysis, although they were retained for denoising, phylogenetic reconstruction, and taxonomic assignment.

@@

(ref:readcounts) Number of Amplicon Sequence Variants (ASV) and reads at different pipeline stages.
*Raw*: raw reads as delivered by sequencing center;
*Trim*: reads with primers and demultiplexing barcodes removed;
*LSUx*: reads with a positive CM hit for 5.8S, allowing regions to be extracted;
*ITS2*: ASVs obtained for the ITS2 region, and reads successfully mapped to an ITS2 ASV;
*short*, *ITS*, *LSU*, *long*: ITS2-based ASVs mapped to consensus ASVs for longer regions, where \"short\" and \"long\" denote the full-length short and long amplicons, respectively.
Short amplicons were mapped to full ITS, LSU, and long amplicon ASVs when they had a perfect match in the ITS2 region.

@@

rdd(bioinf_table) %>%
  set_names(gsub(".*_", "", names(.))) %>%
  kable(
    booktabs = TRUE,
    caption = "(ref:readcounts)",
    align = "rrrrrr",
    format.args = list(big.mark = " "),
    linesep = c(rep("", 3), rep("\\addlinespace", 2), rep("", 4)),
    escape = FALSE
    ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      rdd(bioinf_header)$amplicon[c(TRUE, FALSE)] %>% set_names(rep(2, length(.)), .)
    )
  ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      rle(rdd(bioinf_header)$tech) %$% set_names(lengths, values)
    )
  )
```

@@

@@

```{r venn, fig.cap="(ref:venn)", fig.show="hold", out.width = "90%"}
venn_A <- venndata(rdd(asv_table), ASVs, cols = c("pb_483", "SH-2257", "is_057")) %>%
  vennplot_data(ASVs) %>% {
    plot(
      select(., set, ASVs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
      legend = FALSE,
      labels = FALSE,
      quantities = .$label,
      main = "a",
      fills = c("grey85", "lightblue", "lightcoral"))
  }

@@

grid::grid.legend(
  labels =  c("PacBio Short", "Illumina", "Ion Torrent"),
  nrow = 1,
  pch = 21,
  gp = grid::gpar(fill = c("grey85", "lightblue", "lightcoral")),
  vp = grid::viewport(y = unit(0.1, "npc"))
)

@@

venn_D <- rdd(otu_table) %>%
  mutate_if(is.numeric, ~ . / sum(.)) %>%
  mutate(long = pb_500,
         short = rowMeans(.[,c("pb_483", "SH-2257", "is_057")])) %>%
  select(seq, long, short) %>%
  venndata(OTUs, cols = c("long", "short")) %>%
  vennplot_data(OTUs) %>% {
    plot(
      select(., set, OTUs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
      legend = FALSE,
      labels = FALSE,
      quantities = .$label,
      main = "d",
      fills = c("white", "slateblue1")
    )
  }
# ggpubr::ggarrange(venn_C, venn_D, ncol = 2)
# grid::grid.legend(
#   labels =  c("Long amplicon", "Short amplicon"),
#   nrow = 1,
#   pch = 21,
#   gp = grid::gpar(fill = c("white", "slateblue1")),
#   vp = grid::viewport(y = unit(0.2, "npc"))
# )
```

(ref:readcap) Comparison between read numbers for different sequencing strategies, by ASV (top) and 97% OTU (bottom). ASVs/OTUs which were detected by one sequencing strategy but not the other are plotted as tick marks along the axes. Dashed line represents a constant ratio of read numbers. The blue line is a LOESS smooth of the data, with associated uncertainty in grey shading. $R^2$ value displayed is for log-transformed non-zero read numbers.

```{r pb-vs-it, fig.cap="(ref:readcap)", fig.height=8}
ggpubr::ggarrange(
  read_comparison(
    rdd(multi_table),
    rdd(comparisons),
    type = "ASV"
  ),
  read_comparison(
    rdd(multi_table),
    rdd(comparisons),
    type = "OTU"
  ),
  ncol = 1,
  labels = "auto",
  label.x = 0.9
)
```

@@

Figure\ \@ref(fig:pb-vs-it) shows the correspondence between the read count for different ASVs/OTUs in the different technologies, where shared ASVs/OTUs are plotted as circles, and unshared OTUs are plotted on the margins.
In all cases, the read counts for shared ASVs/OTUs were correlated , with a minimum $R^2$ value of `r compR2("pb_500", "is_057", "ASV")`.
Correlations between read counts for the three technologies using the short amplicon library were increased by OTU clustering
(`r compR2("pb_483", "SH-2257", "ASV")` to `r compR2("pb_483", "SH-2257", "OTU")`,
`r compR2("pb_483", "is_057", "ASV")` to `r compR2("pb_483", "is_057", "OTU")`, and
`r compR2("SH-2257", "is_057", "ASV")` to `r compR2("SH-2257", "is_057", "OTU")`,
for PacBio vs. Illumina, PacBio vs. Ion Torrent, and Illumina vs. Ion Torrent, respectively),
but this was not, or only weakly, the case for correlations between the long amplicon library and short amplicon library
(`r compR2("pb_500", "pb_483", "ASV")` to `r compR2("pb_500", "pb_483", "OTU")`,
`r compR2("pb_500", "SH-2257", "ASV")` to `r compR2("pb_500", "SH-2257", "OTU")`, and
`r compR2("pb_500", "is_057", "ASV")` to `r compR2("pb_500", "is_057", "OTU")`, for PacBio long amplicon reads vs PacBio, Illumina, and Ion Torrent short reads, respectively).

>>>>

tinytex::tlmgr_install("pdfcrop")

knitr::opts_chunk$set(
  dev = "pdf",
  # dev.args = list(pointsize=32),
  echo = FALSE,
  # results = 'asis',
  # fig.show = 'asis',
  fig.width = 6,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  external = TRUE,
  out.width = "70%",
  fig.align = "center"
)
options(
  tikzDefaultEngine = "luatex",
  tikzXelatexPackages = c(options("tikzXelatexPackages")$tikzXelatexPackages,
                          "\\usepackage[version=4]{mhchem}\n",
                          "\\usepackage{siunitx}\n",
                          "\\DeclareSIUnit\\year{a}\n",
                          "\\usepackage{fontspec}"),
  tikzMetricsDictionary = here("writing/transect_paper-tikzDictionary")
)

@@

Ectomycorrhizal (ECM) communities are characterized by mutualistic symbiosis between certain fungi and plant roots.
Spatial turnover rate is one important characteristic of ecological communities, which is relevant for biodiversity assessment and sampling design.
We investigated turnover rate in an ECM fungal community using soil metabarcoding with four different sequencing strategies: short amplicon sequencing of the ITS2 region (300--400\ bp) with Illumina MiSeq, Ion Torrent Ion S5, and PacBio RS II, as well as long amplicon sequencing of the full ITS and partial LSU regions (1200--1600\ bp) with PacBio RS II.
Resulting turnover rates depended more on statistical method than sequencing technology.
We found significant autocorrelation in distance classes <3\ m using a Mantel correlogram, and a best-fit distance-decay curve range of 17--19\ m, both shorter than previously reported for similar habitats.
The use of long-amplicon sequencing enables construction of a phylogenetic tree from metabarcoding reads, which facilitates taxonomic identification of sequences.
However, long reads present issues for denoising algorithms in diverse communities.
We present a solution that splits the reads into shorter homologous regions prior to denoising, and then reconstructs the full denoised reads.
In the choice between short and long amplicons, we suggest a hybrid approach using short amplicons for sampling breadth and depth, and long amplicons to characterize the local species pool for improved identification and phylogenetic analyses.

@@

Several studies have investigated the turnover scale in ECM forests, but so far they have been based on analysis of ECM root tips, rather than bulk soil.
@lilleskov2004 found autocorrelation only at ranges <2.6\ m at most sites in temperate forests, based on Sanger sequencing.
Similarly, @pickles2012a found autocorrelation at distances <3.4\ m, also in temperate forests, based on T-RFLP.
Two studies have included data from tropical African woodlands and wooded savannahs.
@tedersoo2011 found autocorrelation at ranges <10\ m in Miombo woodland based on Sanger sequencing.
However, this study was based on irregular sampling with a minimum spacing of 8\ m, so this as the smallest distance class.
@Bahram2013 performed a meta-analysis of turnover scale in ECM communities, including data from Benin, where they found turnover range of 65\ m based on distance-decay curves.
They concluded that the range of distance decay is greater in tropical forests than temperate forests.

@@

Sampling was conducted at two sites (Ang: N 9.75456° W 2.14064°; Gan: N 9.75678° W 2.31058 °) approximately 30\ km apart in the _Forêt Reservée de l'Ouémé Supérieur_ (Upper Ouémé Forest Reserve) in central Benin.
Both locations were located in woodlands dominated by the ECM host tree _Isoberlinia doka_ (Caesalpinioideae).
At each site, 25\ soil samples were collected along a linear transect at intervals of 1\ m in May 2015.
One third of the sample locations (3\ m spacing) were resampled one year later in June 2016.
For each sample, any coarse organic debris <!--Do you have any definition? Wider than 5 mm? MR--> was removed from the soil surface and a sample of approximately 5cm×5cm×5cm was extracted with a <!-- sterilized? -MR--> knife blade.
Each sample was sealed in a plastic zipper bag and homogenized by shaking and manually breaking apart soil aggregations.
Approximately 50\ mg total of soil was collected from two locations in the homogenized soil sample and placed into a separate 2.0\ mL microtube containing 750\ mL of lysis buffer and lysis beads (Xpedition^TM^ Soil/Fecal DNA miniprep, Zymo Research Corporation, Irvine, California, USA) and lysed in the field using a handheld bead-beater (TerraLyser^TM^; Zymo Research Corporation).
<!--The remainder of each sample was dried at 40°C with an electric dehydrator within 24-48 hours of collection for bulk chemical analysis. -->

@@

After field lysis, DNA was extracted using the Xpedition^TM^ Soil/Fecal Prep kit (see above).
Samples preserved using LifeGuard were first centrifuged at 10000×g for 1\ minute, after which the supernatant was removed and DNA was extracted from the remaining soil using the Soil/Fecal Prep kit as for the other samples.
DNA was quantified using fluorometrically using Quant-iT^TM^ PicoGreen^TM^ dsDNA (Thermo Fisher Scientific, Waltham, MA, USA) fluorescent indicator dye on a Infinite F200 plate spectrofluorometer (Tecan Trading AG, Männedorf, Switzerland) according to the manufacturer's protocol.

Two different fragments of the nuclear rDNA were amplified (Figure\ \@ref(fig:rDNA)).
The short amplicon (approximately 300\ bp) targetted the full ITS2 region as well as parts of the flanking 5.8S and large subunit (LSU) rDNA, using gITS7 [@ihrmark2012] as the forward primer and a mix of ITS4 [@white1990amplification] and ITS4a [@urbina2016] as the reverse primer (hereafter, ITS4m).
The long amplicon (approximately 1500\ bp) targetted the full ITS region including the 5.8S rDNA and approximately 950\ bp at the 5' end of the LSU, including the first three variable regions (Figure\ \@ref(fig:rDNA)), using ITS1 [@white1990amplification] as the forward primer and LR5 [@vilgalys1990] as the reverse primer.

The gITS7 primers for the short amplicon were indexed for multiplexing (**Supplementary File**).
Amplification was performed by polymerase chain reaction (PCR) in 20µl reactions containing 200\ µM dNTP mix, 250\ µM indexed gITS7 primer, 150µM ITS4m, 2mM MgCl~2~, 0.1\ U *Taq* polymerase (Dream *Taq*, Thermo Fisher Scientific, Waltham, MA, USA) and 3--7\ ng purified DNA in Dream *Taq* buffer.
The reaction conditions were 10\ min at 95°, followed by 35\ cycles of 60\ s at 95°, 45\ s at 56°, and 50\ s at 72°, and finally 3\ min at 72°.
Each reaction was conducted in three technical replicates to reduce the effect of PCR stochasticity, which were pooled after amplification.

(ref:rDNA-s) Portion of the rDNA showing the 5.8S rDNA, partial SSU and LSU rDNA, and internally transcribed spacer (ITS) regions

@@

<!--If limited by space, this could be in supp, and then just described in text. -MR-->
```{tikz rDNA, fig.cap = "(ref:rDNA)", fig.scap = "(ref:rDNA-s)", code = readLines("rDNA_amplicons.pgf"), fig.ext = "pdf"}

@@

Both primers for the long amplicon were indexed for multiplexing (**Supplementary File**).
PCR was performed as for the short amplicons, but with 500\ µM of each of the two primers.
Reaction conditions were 10\ min at 95°, 30\ cycles of 45\ s at 95°, 45\ s at 59°, and 90\ s at 72°, and finally 10\ min at 72°.
Each reaction was performed in three technical replicates as for short amplicons.

@@

Amplicons were purified using SPRI beads [@vesterinen2016] and quantified fluorometrically as above.
An aliquot of 100\ ng of DNA from each sample (or the total PCR product if less than 100\ ng) was pooled into two libraries each for long and short amplicons.
Each library was sequenced using Single Molecule Real Time (SMRT) sequencing on a Pacific Biosciences (PacBio) RS II sequencer at the Uppsala Genome Center (UGC; Uppsala Genome Center, Science
for Life Laboratory, Dept. of Immunology, Genetics and Pathology, Uppsala University, BMC, Box 815, SE-752\ 37 UPPSALA, Sweden).
Short amplicon libraries were sequenced on two SMRT cells each, while long amplicon libraries were sequenced on four SMRT cells each.

Additionally the short amplicon libraries were combined and sequenced using an Ion S5 (Ion Torrent) sequencer using one 520 chip at UGC, and a MiSeq (Illumina Inc.) sequencer using v3 chemistry with a paired-end read length of 300\ bp at the SNP&SEQ Technology Platform (Dept. of Medical Sciences, Uppsala University, BMC, Box 1432, SE-751\ 44 UPPSALA, Sweden).
The Illumina library was pooled with samples for another project, with one half of the reads from one lane devoted to this project.

@@

Raw reads were divided into shorter regions by matching to covariance models (CM), which are similar to stochastic hidden markov models (HMM), but account for both nucleotide sequence and RNA secondary structure [@eddy1994].
First, the 5.8S rDNA was located in each read by searching for Rfam model RF0002 [@kalvari2018] using `cmsearch` from Infernal 1.1.2 [@nawrocki2013], and all bases before the 5.8S were assigned to ITS1.
No attempt was made to remove the approximately 12\ bp fragment of the SSU from the 5' end of ITS1 in the long amplicons; it was too short to be reliably detected by a CM or the HMMs employed by ITSx [@bengtsson-palme2013].
<!--This approach was able to identify both the 5' and 3' ends of the 5.8S rDNA more consistently and quickly than ITSx [@bengtsson-palme2013, data not shown].-->
A reference alignment including conserved RNA base pairing between and within the 5.8S and relevant portions of LSU was generated from the fungal 28S RNA seed alignment from the Ribosomal Data Project (RDP) release 11.5 [@glockner2017; @cole2014] by truncating after the LR5 primer site and using the reference line to annotate the variable regions *sensu* @michot1984 and @raue1988.
A CM was generated from the alignment using `cmbuild` from Infernal.
The fragment of each read beginning with the 5.8S rDNA was then aligned to the CM using `cmalign` from Infernal.
The annotation line in the CM alignment for each read was then used to split the reads into alternating more-conserved and less-conserved regions as shown in Figure\ \@ref(fig:rDNA),
where LSU1-4 represent the conserved regions of LSU flanking the variable D1-3 regions [@michot1984].
For short amplicons, only (partial) 5.8S, ITS2, and (partial) LSU1 were extracted.
Code to extract the regions, including annotated seed alignments and CMs, is available in the new R package `LSUx`, available on github (https://github.com/brendanf/LSUx).

@@

Full length long amplicon ASVs were aligned using DECIPHER [@wright2015] with up to 10\ iterations of progressive alignment and conserved RNA secondary structure calculation and 10\ refinement iterations.
This alignment was truncated at a position after the D3 region corresponding to base 907 of the *Saccharomyces cerevisiae* S288C reference sequence for LSU, because several sequences had introns
<!-- Just curious, what are Type 1 introns -AR--> <!-- It has to do with the enzyme that cuts them out, I think, but I don't actually know my introns are Type 1, just that this is a location where Type 1 introns are known to occur. -BF-->
after this position, as also observed in several fungal species by @holst-jensen1999.

@@

Ecological community dissimilarity matrices were calculated using the ASV/OTU based Bray-Curtis metric (both long and short amplicons) and the phylogenetically based weighted UNIFRAC method (only long amplicons) in `phyloseq` version `r packageVersion("phyloseq")`.
Each of these distance matrices was used to calculate a Mantel correlogram for distances of 0--12\ m.
Separate correlograms were drawn for samples taken during the same year, and samples separated in time by one year, in order to assess the degree to which the soil community changes over the course of one year.

@@

DNA concentrations after extraction and PCR, as well as sequencing reads for PacBio and Illumina, are shown per sample in Supplementary Figure \@ref(fig:sample-depth).
Samples from Ang in 2015 yielded low quantities of DNA, poor PCR performance, and ultimately very few sequencing reads, especially in the long amplicon library, where only one sample produced more than 100\ reads.
Consequently Ang samples were excluded from spatial analysis, although they were retained for denoising, phylogenetic reconstruction, and taxonomic assignment.

@@

(ref:readcounts-s) Number of Amplicon Sequence Variants (ASV) and reads at different pipeline stages

(ref:readcounts) *Raw*: raw reads as delivered by sequencing center;
*Trim*: reads with primers and demultiplexing barcodes removed;
*LSUx*: reads with a positive CM hit for 5.8S, allowing regions to be extracted;
*ITS2*: ASVs obtained for the ITS2 region, and reads successfully mapped to an ITS2 ASV;
*short*, *ITS*, *LSU*, *long*: ITS2-based ASVs mapped to consensus ASVs for longer regions, where \"short\" and \"long\" denote the full-length short and long amplicons, respectively.
Short amplicons were mapped to full ITS, LSU, and long amplicon ASVs when they had a perfect match in the ITS2 region.

@@

rdd(bioinf_table) %>%
  set_names(gsub(".*_", "", names(.))) %>%
  kable(
    booktabs = TRUE,
    caption = "(ref:readcounts-s). (ref:readcounts)",
    caption.short = "(ref:readcounts-s)",
    align = "rrrrrr",
    format.args = list(big.mark = " "),
    linesep = c(rep("", 3), rep("\\addlinespace", 2), rep("", 4)),
    escape = FALSE
    ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      rdd(bioinf_header)$amplicon[c(TRUE, FALSE)] %>% set_names(rep(2, length(.)), .)
    )
  ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      rle(rdd(bioinf_header)$tech) %$% set_names(lengths, values)
    )
  )
```

@@

(ref:venn-s) Venn diagrams of shared ASVs and OTUs between different sequencing technologies

@@

```{r venn, fig.cap="(ref:venn)", fig.scap = "(ref:venn-s)", fig.show="hold", out.width = "70%", fig.height = 4, fig.align="center", results="asis"}
venn_A <- venndata(rdd(asv_table), ASVs, cols = c("pb_483", "SH-2257", "is_057")) %>%
  vennplot_data(ASVs) %>% {
    plot(
      select(., set, ASVs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
      legend = FALSE,
      labels = FALSE,
      quantities = .$label,
      main = "a",
      fills = c("grey85", "lightblue", "lightcoral"))
  }

@@

grid::grid.legend(
  labels =  c("PacBio Short", "Illumina", "Ion Torrent"),
  nrow = 1,
  pch = 21,
  gp = grid::gpar(fill = c("grey85", "lightblue", "lightcoral")),
  vp = grid::viewport(y = unit(0.05, "npc"))
)

@@

venn_D <- rdd(otu_table) %>%
  mutate_if(is.numeric, ~ . / sum(.)) %>%
  mutate(long = pb_500,
         short = rowMeans(.[,c("pb_483", "SH-2257", "is_057")])) %>%
  select(seq, long, short) %>%
  venndata(OTUs, cols = c("long", "short")) %>%
  vennplot_data(OTUs) %>% {
    plot(
      select(., set, OTUs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
      legend = FALSE,
      labels = FALSE,
      quantities = .$label,
      main = "d",
      fills = c("white", "slateblue1")
    )
  }
ggpubr::ggarrange(venn_C, venn_D, ncol = 2)
grid::grid.legend(
  labels =  c("Long amplicon", "Short amplicon"),
  nrow = 1,
  pch = 21,
  gp = grid::gpar(fill = c("white", "slateblue1")),
  vp = grid::viewport(y = unit(0.15, "npc"))
)
```

(ref:read-compare-s) Comparison between read numbers for different sequencing strategies

(ref:read-compare) by ASV (top) and 97% OTU (bottom). ASVs/OTUs which were detected by one sequencing strategy but not the other are plotted as tick marks along the axes. Dashed line represents a constant ratio of read numbers. The blue line is a LOESS smooth of the data, with associated uncertainty in grey shading. $R^2$ value displayed is for log-transformed non-zero read numbers.

```{r read-compare, fig.cap="(ref:read-compare-s), (ref:read-compare)", fig.scap="(ref:read-compare-s)", fig.height=8}
ggpubr::ggarrange(
  read_comparison(
    rdd(multi_table),
    rdd(comparisons),
    type = "ASV"
  ),
  read_comparison(
    rdd(multi_table),
    rdd(comparisons),
    type = "OTU"
  ),
  ncol = 1,
  labels = "auto",
  label.x = 0.9
)
```

@@

Figure\ \@ref(fig:read-compare) shows the correspondence between the read count for different ASVs/OTUs in the different technologies, where shared ASVs/OTUs are plotted as circles, and unshared OTUs are plotted on the margins.
In all cases, the read counts for shared ASVs/OTUs were correlated , with a minimum $R^2$ value of `r compR2("pb_500", "is_057", "ASV")`.
Correlations between read counts for the three technologies using the short amplicon library were increased by OTU clustering
(`r compR2("pb_483", "SH-2257", "ASV")` to `r compR2("pb_483", "SH-2257", "OTU")`,
`r compR2("pb_483", "is_057", "ASV")` to `r compR2("pb_483", "is_057", "OTU")`, and
`r compR2("SH-2257", "is_057", "ASV")` to `r compR2("SH-2257", "is_057", "OTU")`,
for PacBio vs. Illumina, PacBio vs. Ion Torrent, and Illumina vs. Ion Torrent, respectively),
but this was not, or only weakly, the case for correlations between the long amplicon library and short amplicon library
(`r compR2("pb_500", "pb_483", "ASV")` to `r compR2("pb_500", "pb_483", "OTU")`,
`r compR2("pb_500", "SH-2257", "ASV")` to `r compR2("pb_500", "SH-2257", "OTU")`, and
`r compR2("pb_500", "is_057", "ASV")` to `r compR2("pb_500", "is_057", "OTU")`, for PacBio long amplicon reads vs PacBio, Illumina, and Ion Torrent short reads, respectively).

Almost all of the short amplicons from all three technologies were between 240 and 375\ bp long (Supplementary Figure\ \@ref(fig:full-length)a).
Although the length profile of the three requencing runs were very similar, Illumina MiSeq had the largest fraction of reads near the top of the range, followed by Ion Torrent Ion S5 and PacBio RS II (Supplementary Figure\ \@ref(fig:full-length)b).
In contrast, the length of the long amplicon reads varied widely, from `r filter(rdd(reads_table), region == "long") %$% range(length) %>% paste(collapse = " to ")` bp, with a median of `r filter(rdd(reads_table), region == "long") %$% reldist::wtd.quantile(length, weight = reads)` bp.
